# Analysis of Longitudinal Data
Today is
```{r echo=FALSE}
date()
```

This week we practise the analysis of longitudinal data. We will make graphical displays and summaries. We use the data RATS given by here: \url{https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt} The data is from a nutrition study conducted in three groups of rats (Crowder and Hand, 1990). The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately
weekly, except in week seven when two recordings were taken) over a 9-week
period.

```{r}
# Access the packages dplyr and tidyr
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)

# Read the data
RATSL <- read.csv("/Users/teemupekkarinen/IODS-project/data/RATSL",row.names = 1,stringsAsFactors = T)
str(RATSL)
dim(RATSL)
summary(RATSL)

# Factor ID & Group
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# Take a glimpse at the RATSL data
glimpse(RATSL)

# Table 
kable(RATSL[1:16,])

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Rats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Rats), max(RATSL$Rats)))

```

That is, there are different 16 rats for which the table above gives the initial weight in day 1. The first figure shows how the weight of rats in each group evolve over the test period. Clearly there are differences in weights between groups. Moreover, in group 2 there is a clear outlier of a huge rat.
```{r}

# Standardise the variable Rats
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdRats = (Rats - mean(Rats))/sd(Rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised Rats
ggplot(RATSL, aes(x = Time, y = stdRats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized Rats")

```

```{r}

# Number of time periods, baseline (time 1) included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of Rats by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Rats), se = sd(Rats)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype=Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=4) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.5)) +
  scale_y_continuous(name = "mean(Rats) +/- se(Rats)")

```

```{r}

# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline time period 1).
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Rats), time periods 2-64")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- RATSL8S %>%
  filter(mean < 550)

# Glimpse the data
glimpse(RATSL8S1)

# Draw a boxplot of the mean versus treatment
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Rats), time periods 2-64")

```