
# About the Project

Introduction to Open Data Science. Happy to learn to connect Git and R even though the content of data analysis is *familiar* to me. I've used R MarkDown before and found it useful. In addition I've used Git before on [Tietokantojen Perusteet](https://github.com/teemupek/Tikape_Chat.git) course. However, it's been a long time so recalling things is needed. Here's the [link](https://github.com/teemupek/IODS-project.git) to my IODS GitHub repository.

Today is the following day.
```{r echo=FALSE}
# This is a so-called "R chunk" where you can write R code.

date()

```

## Reminders

I wanted to remind me what I've done last time with R MarkDown. I found a nice data on **exchange and forward rates**. I make a table and a graph of Sterling/EUR exchange rate. I uploaded `forward2.dat` to git repository and I call my data set from there in order to plot the rate. I hide the R code to make the outcome more readable.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, include=FALSE}

 #### Packages ####
rm(list=ls()) 
library(dplyr)
library(data.table)
library(lmtest)
library(tidyr)
library(sandwich)
library(knitr)

```

### Table of the exchange and forward rates
```{r echo=FALSE, results='asis'}

#### Data ####

# EXUSBP : exchange rate USDollar/British Pound Sterling
# EXUSEUR: exchange rate USDollar/euro
# EXEURBP: exchange rate euro/pound
# F1USBP : 1 month forward rate USD/pound
# F1USEUR: 1 month forward rate USD/euro
# F1EURBP: 1 month forward rate euro/pound
# F3USBP : 3 month forward rate USD/pound
# F3USEUR: 3 month forward rate USD/euro
# F3EURBP: 3 month forward rate euro/pound
forward2 <- fread("forward2.dat")
forward2 <- as_data_frame(forward2)
kable(summary(forward2))

# Let's logarithmised the data
fwd <- log(forward2)

# Colnames
colnames(fwd) <- c("s_usd_gbp",
                   "s_usd_eur",
                   "s_eur_gbp",
                   "f1_usd_gbp",
                   "f1_usd_eur",
                   "f1_eur_gbp",
                   "f3_usd_gbp",
                   "f3_usd_eur",
                   "f3_eur_gbp")

```

### Graph of the GBP/EUR exchange rate.

```{r echo=FALSE}

tsEurBp <- ts(exp(fwd$s_eur_gbp), frequency=12, start=c(1979,1), end = c(2001,12)) 
plot(tsEurBp, main="GBP/EUR", xlab="", ylab="")

```

## Latex
I'm happy to see that R MarkDown is linked to LaTeX syntax! Here's a maximization problem from my current research on the optimal mechanism design with enforcement:
\begin{align}
		\max_{r(\cdot),t(\cdot),m(\cdot)} \mathbb{E} \left[ t(\theta) - K m(\theta) \right]\label{OBJ} \tag{MAX}
	\end{align}
	subject to
	\begin{align}
		&t(\theta) = \theta r(\theta) - V(\underline{\theta}) - \int_{\underline{\theta}}^{\theta} \mathcal{I}(s|s)ds \label{TAX} \tag{TAX} \\
		&\mathcal{I}(\theta|\theta) \qquad \text{is nondecreasing} \label{IC} \tag{IC} \\
		&\mathcal{I}(\theta|\theta) \geq 0 \label{IR} \tag{IR} 
	\end{align}
	for all $\theta \in \Theta$ with $\mathcal{I}(\theta|\theta):=r(\theta) - m(\theta) \varphi$.